\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Local Adaptive Optimization of Time Step}

### 2013-10-27 report/article.tex
Methods described

\subsection{1D mass-spring system}
Mass-spring systems are systems simulating a set of vertices interconnected by
springs in various ways to obtain desirable physical properties. The motion of
the springs and vertices are then simulated as a function of the time elapsed.
Examples of application of mass-spring systems are simulation of cloth or hair.

@@

\subsection{Switch}
Change timestep when detecting collision/switch in next iteration

\subsection{Spatial LAOTS}
Atomic update

Fixed update rate and detection of need for re-calculation of individual springs

Heuristic: Length of spring, steps since last update

\subsection{Inverse}
Time-local optimization according to heuristic:

\begin{align}
    \frac{1}{a} &= \frac{1}{\frac{-kx}{m}} = \frac{m}{-kx} \\
    \Delta t &= \abs{\frac{1}{a}} + \frac{1}{0.1}
\end{align}

@@

\section{Experiments and results}

>>>>

\subsection{1D mass-spring system} Mass-spring systems are systems simulating
a set of vertices interconnected by springs in various ways to obtain
desirable physical properties. The motion of the springs and vertices are
then simulated as a function of the time elapsed. Examples of application of
mass-spring systems are simulation of cloth or hair. When performing simulations
we want to avoid a switch of position of spring-connected vertices caused by the
timestep size of the simulation, since this invalidates our discretization of
the domain. Furthermore we want to perform as few timesteps as possible in order
to get to the final simulation state by using as big timesteps as possible
without affecting the outcome of the simulation.

@@

\subsection{Adapt-on-switch}
The first LAOT method we propose is time-adaptive. We perform a simulation
with an initial timestep size. In each step we try to predict whether a timestep
of the same size using the current velocities will result in a switch of the
left and right vertices of any of the springs. If this is the case, we should
try to avoid the switch by decreasing the timestep of the current iteration,
re-calculate the velocity and position changes and resume stepping of the
simulation using the new timestep size. When switches aren't detected anymore
we gradually increase the timestep of the simulation again. We call this
method the adapt-on-switch method.

The switch of two endspoints connected by a spring $d$ can easily be
calculated by checking the length ($p_{l,d} - p_{r,d}$) of the spring before
and after a position update. If the sign of the length differs, the two
points must have switched sides and hence adaption of the timestep should be
performed. The adaption of the timestep can either be done by division of the
current timestep by a constant or by having a finite set of timestep sizes to
choose from.

\subsection{Atomic updating using heuristics}
The second LAOT method we propose is space-adaptive. We perform simulations
using a fixed timestep size. In each iteration we perform position update of
all vertices, but only re-calculate acceleration and velocity of the springs
and points that fulfill one of the defined heuristics. The heuristics can be
chosen to fit the domain of the simulation, but we have chosen to use the
number of steps since last update and the ratio between the current spring
length and the spring length at the last update. This should force a bound on
the maximal timestep before update as well as making sure updates are done
more frequently in case the springlength is changing more rapidly. Further
details are explained in section \ref{sec:experiments}.

\subsection{Global estimated acceleration}
The third and last LAOT method we propose is time-adaptive. The simulations are
done using an initial timestep size. We know that the force and hence
acceleration of a spring is somewhat linear close to the spring's rest length
and it ``bends'' off as the spring is streched or contracted further away from
the rest length. Based on this knowledge we will try to make a heuristic
telling us about the needed timestep size at any given point during the
simulation. The smallest needed timestep defines the global timestep for that
given step.

For this simple 1D example we use the inverse of the acceleration
for adapting the timestep and hence the heuristic is defined as:
\begin{align}
    \frac{1}{a} &= \frac{1}{\frac{-kx}{m}} = \frac{m}{-kx} \\
    \Delta t &= \text{min}\left( \frac{1}{\abs{a} + 0.1} \right )
\end{align}
We have added a small value (0.1) in order to avoid the timestep to increase
exponentially towards infinity as $\abs{a}$ approach 0. Furthermore a scaling of the timestep
could be needed depending on the scale of the simulation mesh. Further details
are explained in section \ref{sec:experiments}.

Our though is that for some more
complex simulations we are able to make estimates of this type less costly than
performing the actual simulations of a small fixed size timestep.

@@

\section{Experiments and results}
\label{sec:experiments}

Show results for:

\begin{itemize}
    \item Normal simulation
    \item Switch
    \item Atomic
    \item Inverse
\end{itemize}
